---
title: "data-cleaning-exercise"
author: "Kaitlyn"
date: "2/5/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
```

Load escapement goals, get rid of extraneous columns, filter to Bristol Bay region and Sockeye species. Lower and Initial.Year are being read as character, because one row has a Lower of "eliminated" (should be NA).
```{r}
escapement_goals <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/knb.92014.1",
                             stringsAsFactors = FALSE) %>% # read in the file
  filter(Region == "Bristol Bay", Species == "Sockeye") %>% # filter as instructed
  mutate(Lower = ifelse(Lower == "eliminated", NA, Lower)) %>% # replace "eliminated" with "NA"
  mutate(Lower = as.integer(Lower)) %>%  # I suppose this would have done the above for me?
  mutate(Initial.Year = as.integer(Initial.Year)) %>% # convert Initial.Year from character to integer
  select(System, Lower, Upper, Initial.Year) %>% # take only the columns we need
  rename(Location = System) # rename 'System' as 'Location' to match below (and it makes more sense to me)

head(escapement_goals)
```

Load escapement counts. Apply same filters, plus restrict to Locations that are in the goals tab. Also split up date into constituent parts.
```{r}
escapement_counts <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e", stringsAsFactors = FALSE) %>% # read in the file
  rename(Region = SASAP.Region) %>% # rename Region so it matches below
  filter(Region == "Bristol Bay",
         Species == "Sockeye",
         Location %in% escapement_goals$Location) %>% # take only locations that have goals
  separate(col = sampleDate, into = c("Year", "Month", "Day"), sep = "-", remove = FALSE) # split up date

head(escapement_counts)
```

Calculate yearly escapement for each location and join with goals, drop any rows where there is no goal for that year, and determine whether each year met escapement goals.
```{r}
escapement_yearly <- escapement_counts %>% 
  group_by(Year, Location) %>% # group by year and location
  summarise(Total = sum(DailyCount)) 

escapement_joined <- escapement_yearly %>% 
  left_join(escapement_goals, by = "Location") %>% # append the goals to yearly with a left_join
  filter(Initial.Year <= Year) %>% # remove any rows for which the year was BEFORE the goal was set
  mutate(Goal.Met = ifelse((Total > Lower & Total < Upper), 1, 0)) # determine whether count was within goal range

head(escapement_yearly)

#
escapement_goal_outcome <- escapement_yearly %>% 
  group_by(Location) %>% 


head(escapement_goal_outcome)

```


