---
title: "Data Cleaning"
author: "Kaitlyn"
date: "2/4/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read in the data

```{r}
library(dplyr)
library(tidyr)
```

```{r}
catch_original <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", stringsAsFactors = FALSE)
```


## Tidy data

Remove the 'notes' and 'all' columns because we don't need them.

```{r}
catch_clean <- catch_original %>%
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_clean)
```

Alternatively, instead of specifying columns you want, can specify the ones you don't want with the minus sign.

```{r}
catch_clean <- catch_original %>%
  select(-All, -notesRegCode)

head(catch_clean)
```

You can see that something is not right. The Chinook column is a character, not numeric. Let's try to convert to an integer.

```{r}
catch_clean <- catch_original %>%
  select(-All, -notesRegCode) %>%
  mutate(Chinook = as.numeric(Chinook))
```

This code runs, but we have a warning that `NAs introduced by coercion`. Something wasn't able to be converted to a number, so it became an NA.

Determine which values were converted to NA by `as.numeric()`

```{r}
i <- which(is.na(catch_clean$Chinook))
i # can see there is one row
catch_original[i,] 
```

The Chinook column has an "I" instead of a "1." Was a scanned PDF and there was an error in reading. 

So let's actually change how we clean the data. We can use `ifelse()`, a function that does a test (first argument), then returns values based on yes (second argument) or no (third argument).

```{r}
catch_clean <- catch_original %>%
  select(-All, -notesRegCode) %>%
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook))

summary(catch_clean)
```

Chinook is still a character variable; we need to change it. Add back in the `as.numeric` function...

```{r}
catch_clean <- catch_original %>%
  select(-All, -notesRegCode) %>%
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>%
  mutate(Chinook = as.numeric(Chinook))

summary(catch_clean)
```

Now we are in business.

## Convert to long format

Use new tidyverse function `pivot_longer`. 
Note that we are selecting all of the fish columns (everything BUT region, year).
```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year), 
               names_to = "species",
               values_to = "catch")

head(catch_long)
```

We can use `pivot_wider` to reverse what we just did.

```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = "species",
              values_from = "catch")

head(catch_wide)
```

Taking a closer look at the data, we can see that the catch is actually in 1,000s of fish. Let's rename the catch variable to clarify.

```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year), 
               names_to = "species",
               values_to = "catch") %>% 
  rename(catch_thousands = catch)

head(catch_long)
```

And if we want a column that has actual catch, we can rename, then mutate.

```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year), 
               names_to = "species",
               values_to = "catch") %>% 
  rename(catch_thousands = catch) %>% 
  mutate(catch = catch_thousands * 1000)

head(catch_long)
```

And we can get rid of the catch_thousands. We could have done this more cleanly if we had caught it earlier.

```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year), 
               names_to = "species",
               values_to = "catch") %>% 
  rename(catch_thousands = catch) %>% 
  mutate(catch = catch_thousands * 1000) %>%
  select(-catch_thousands)

head(catch_long)
```

Calculate mean catch per region - first group, then summarise

```{r}
mean_region <- catch_long %>% 
  group_by(Region) %>%
  summarise(mean_catch = mean(catch))

head(mean_region)
```

Filter for only sockeye species. Filter is a helpful way to select certain rows.

```{r}
sockeye_catch <- catch_long %>%
  filter(species == "Sockeye")

head(sockeye_catch)
```

Exercise: what region has highest mean sockeye catch during this period.

```{r}
sockeye_mean <- sockeye_catch %>%
  group_by(Region) %>%
  summarise(mean_catch = mean(catch))

sockeye_mean
```

What if we wanted to order them?

Arrange will order them by a given column, in increasing order. If you want them decreasing, can add a minus sign, OR use `desc()` function, which is better for other formats too (not just number, but alphabetical).

```{r}
sockeye_mean <- sockeye_catch %>%
  group_by(Region) %>%
  summarise(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(sockeye_mean)
```

BRB had the highest count.

Now that we have our cleaned data, we want to join in with more information on the regions from the region table.

```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1", stringsAsFactors = FALSE) %>%
  select(code, mgmtArea)

head(region_defs)
```

Let's join the two tables together. You can of course use the pipe, but it's nice to know what is on the left and what is on the right.

```{r, eval = F}
catch_joined <- left_join(catch_long, region_defs)
```

We get an error; there are no common variables between the data sources, because they have different names. It's generally always better to explicitly specify what columns you are joining over.

```{r}
catch_joined <- left_join(catch_long, region_defs,
                          by = c("Region" = "code"))

head(catch_joined)
```

Alternatively, we could have renamed one of the columns first before we joined them.

## Separate and Unite

Let's make a new dataframe to play with.

```{r}
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAI-039"),
                       stringsAsFactors = FALSE)
```

Here, we have a column that contains both island and site number, but we want these to be in separate columns. We can use separate to return two columns, and specify the separator.

```{r}
sites_df %>% 
  separate(col = site, 
           into = c("island", "site_number"), 
           sep = "-")
```

If we wanted to keep that original column, can set `remove = FALSE`

```{r}
sites_df %>% 
  separate(col = site, 
           into = c("island", "site_number"), 
           sep = "-",
           remove = FALSE)
```

The reverse of `separate()` is `unite()`.

Copy in another dummy data frame.

```{r}
dates_df <- data.frame(year = c("1930",
                                "1930",
                                "1930"),
                       month = c("12",
                                "12",
                                "12"),
                       day = c("14",
                               "15",
                               "16"),
                       stringsAsFactors = FALSE)
```

```{r}
dates_df %>% 
  unite(date, year, month, day, sep = "-", remove = FALSE)
```

